Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-26 tex/classical.tex

In Boltzmann kinetic theory of gases, one starts by defining the distribution function, 
$f(\vec{r},\vec{v},t)$, as the average number of molecules that, at time $t$, have 
position $\vec r$ and velocity $\vec v$, and are contained in a $\mu$-space volume element $d^3rd^3v$.
Assuming a deterministic Newtonian description of the molecules motion, 
as well as the invariance of the $\mu$-space volume measure, one arrives to the Boltzmann rate equation:
%
\begin{equation}
	f(\vec{r}+\vec{v}\delta t, \vec{v}+\vec{F} \delta t, t+\delta t)=f(\vec{r},\vec{v},t)+\left( \frac{\partial f}{\partial t} \right)_{\textrm{coll}} \delta t.
\end{equation}
%
Here, the term $\left(\partial f/\partial t\right)_{\textrm{coll}}$
describes the in and out fluxes from and towards the volume element, due to the collisions.
Subsequently, from the previous equation, the integro-differential Boltzmann transport equation is obtained:
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the $H$-functional \cite{bib:tolman,bib:huang}:
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Notice that $f_B(\vec{v},t)$ is a spatially homogeneous distribution function.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a diluted gas occupying a volume $V$, at temperature $T$, with total energy $E$, and total
number of free classical particles $N$.
In order to clearly distinguish the Boltzmann functional $H_B$, we denote hereafter
the Maxwell-Boltzmann distribution function as $f_B$.

The physically correct espontaneous time-evolution of an out of equilibrium diluted gas 
is corroborated by the $H$-theorem.
This theorem establishes that if the two conditions, a) the homogeneous
function $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and b) the molecular chaos hypothesis is valid, then
the system evolves in such a manner that $dH_B/dt\leq0$, and if $dH_B/dt=0$, 
then the system is in the equilibrium state. 
The $H$-theorem is straightforward to prove through Eq.~(\ref{eq:transport}) \cite{bib:huang}, 
and it assures the consistency between our microscopic approach to describe the system's spontaneous 
time-evolution and the phenomenological observations
established by the $2^{\textrm{nd}}$ law of classical thermodynamics; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a diluted gas in equilibrium
with no applied external forces, \textit{i.e.}
$(\partial f/\partial t)=0$ and $f$ is independent of $\vec r$), 
it can be directly proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm of Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposed $H$-functional, we remark the following. In defining
$H_B$, Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function 
$f$ is spatially homogeneous. This assumption simplifies the demostration 
of the $H$ theorem, however, it introduces as well a limited conception
of the out-of-equilibrium condition of the gas. Given the relatively simple nature of a dilute 
gas, one of the salient features of an out-of-equilibrium condition is the existence of
inhomogeneities in the system.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is fullfiled. In order to extend the previous procedure to systems whose
distribution functions are non-homogeneous, which might allow for studying systems
being in a more general out-of-equilibrium condition, we introduce a modified $H$-functional 
in the following. For the sake of clearity and simplicity, we will use primed functions
and quantities to denote the classical case, so as to differentiate them from the quantum
analogues.

@@

We now use the variational method to find the extremal of $\mathcal{H}'$,
consistent with the restrictions
(\ref{eq:cellrestrictions}) and (\ref{eq:micro}), together with the corresponding Lagrange
multipliers $\{\alpha_M\}$ and $\{\beta_M\}$. This yields:
%
\begin{eqnarray}\label{eq:deltaHpdeltafpj}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that the use of the variational procedure on
$\mathcal{H}'$ predicts that when the equilibrium is reached, the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

Furthermore, in equilibrium, the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
Since $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$ (a constant)
and because we are mainly interested in the time-derivative of $\mathcal{H}''$,
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ leads to the Maxwell-Boltzmann
distribution function of systems in equilibrium.

>>>>

In Boltzmann kinetic theory of gases, one starts by defining the distribution function, 
$f(\vec{r},\vec{v},t)$, as the average number of molecules that, at time $t$, have 
position $\vec r$ and velocity $\vec v$, and are contained in a $\mu$-space volume element $d^3rd^3v$.
Assuming a deterministic Newtonian description of the molecules' motion, 
as well as the invariance of the $\mu$-space volume measure, one arrives to the Boltzmann rate equation:
%
\begin{equation}
	f(\vec{r}+\vec{v}\delta t, \vec{v}+\vec{F} \delta t, t+\delta t)=f(\vec{r},\vec{v},t)+\left( \frac{\partial f}{\partial t} \right)_{\textrm{coll}} \delta t.
\end{equation}
%
Here, the term $\left(\partial f/\partial t\right)_{\textrm{coll}}$
describes the in and out fluxes from and towards the volume element, due to the collisions.
Subsequently, from the previous equation, the integro-differential Boltzmann transport equation is obtained:
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

The evolution of a dilute gas towards thermodynamic equilibrium is frequently
addressed by first defining the $H$-functional \cite{bib:tolman,bib:huang}:
%
\begin{equation}\label{eq:hbfunctional}
   H_{B}=\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v.
\end{equation} 
%
Notice that $f_B(\vec{v},t)$ is a spatially homogeneous distribution function.
The functional $H_B$,
originally introduced by Boltzmann in 1872,
describes a dilute gas occupying a volume $V$, at temperature $T$, with total energy $E$, and total
number of free classical particles $N$.
In order to clearly distinguish the Boltzmann functional $H_B$, we denote hereafter
the Maxwell-Boltzmann distribution function as $f_B$.

The physically correct espontaneous time-evolution of an out of equilibrium dilute gas 
is corroborated by the $H$-theorem.
This theorem establishes that if the two conditions, a) the homogeneous
function $f(\vec{v},t)$ satisfies the Boltzmann
transport equation and b) the molecular chaos hypothesis is valid, then
the system evolves in such a manner that $dH_B/dt\leq0$, and if $dH_B/dt=0$, 
then the system is in the equilibrium state. 
The $H$-theorem is straightforward to prove through Eq.~(\ref{eq:transport}) \cite{bib:huang}, 
and it assures the consistency between our microscopic approach to describe the system's spontaneous 
time-evolution and the phenomenological observations
established by the $2^{\textrm{nd}}$ law of classical thermodynamics; in fact,
$H_B$ can be associated with an entropy density.

On the other hand, considering a dilute gas in equilibrium
with no applied external forces, \textit{i.e.}
$(\partial f/\partial t)=0$ and $f$ is independent of $\vec r$, 
it can be directly proven that the equilibrium distribution function obtained from
Eq.~(\ref{eq:transport}) is precisely the Maxwell-Boltzmann
distribution function. The proof of the above commences by finding the sufficient
condition for $f$ to render a null r.h.s. of Eq. (\ref{eq:transport}). Such an
$f$, which we denote here as $f_0$, must satisfy:
%
\begin{equation}\label{eq:neccondf0}
    f_0(\vec{v}'_2)f_0(\vec{v}'_1)- f_0(\vec{v}_2)f_0(\vec{v}_1)=0.
\end{equation}
%
Subsequently, the Maxwell-Boltzmann distrubution function can be obtained by
taking the logarithm of Eq.~(\ref{eq:neccondf0}), and exploiting mechanical
conserved quantities (see \cite[ch. 4.2]{bib:huang}).

Before introducing our proposed $H$-functional, we remark the following. In defining
$H_B$, Eq.~(\ref{eq:hbfunctional}), it is assumed that the distribution function 
$f$ is spatially homogeneous. This assumption simplifies the demostration 
of the $H$ theorem, however, it introduces as well a limited conception
of the out-of-equilibrium condition of the gas. Given the relatively simple nature of a dilute 
gas, one of the salient features of an out-of-equilibrium condition is the existence of
inhomogeneities in the system, which is not considered in the above.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is fulfilled. In order to extend the previous procedure to systems whose
distribution functions are non-homogeneous, which might allow for studying systems
being in a more general out-of-equilibrium condition, we introduce a modified $H$-functional 
in the following. For the sake of clarity and simplicity, we will use primed functions
and quantities to denote the classical case, so as to differentiate them from the quantum
analogues.

@@

We now use the variational method to find the extremal of $\mathcal{H}'$,
consistent with the restrictions
(\ref{eq:cellrestrictions}) and (\ref{eq:micro}), together with the corresponding Lagrange
multipliers $\{\alpha_M\}$ and $\{\beta_M\}$. This yields:
%
\begin{eqnarray}\label{eq:deltaHpdeltafpj}
    \frac{\delta \mathcal{H}'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that applying the variational procedure on
$\mathcal{H}'$ predicts that when the equilibrium is reached, the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

Furthermore, in equilibrium, the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $\mathcal{H}''$
functional:
%
\begin{equation}\label{CH3}
   \mathcal{H}''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v.
\end{equation}
%
Since $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$ (a constant)
and because we are mainly interested in the time-derivative of $\mathcal{H}''$,
$C$ can be conveniently omitted. In other words, $\mathcal{H}'$ leads to the Maxwell-Boltzmann
distribution function of systems in equilibrium.

### 2021-01-26 tex/disscussion.tex

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions both in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. \commb{The relaxation process of the system is described by
monotonic functions that account for deviations from the global equilibrium.}

@@

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equlibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of diluted gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of out-of-equilibrium systems, other than diluted gases. In particular, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the functional form of $g'_M$ and $g_{nM}$.

>>>>

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions both in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. The relaxation process of the system is described by
monotonic functions that account for deviations from the global equilibrium.

@@

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equlibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of dilute gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of out-of-equilibrium systems, other than dilute gases. In particular, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the explicit functional form of $g'_M$ and $g_{nM}$.

### 2021-01-26 tex/qccorrespondence.tex

In sections \ref{sec:classical} and \ref{sec:quantum}, we have seen that the variational
method can be applied to $H$-functionals, which correctly describes the behavior of
classical and quantum diluted gases, as to their respective time-evolution.
Both $H$-functionals defined in Eqs.~(\ref{eq:cHdef}) and
(\ref{eq:qHdef}) also recovers the well-known distribution functions, either Maxwell-Boltzmann
for a classical gas, Fermi-Dirac for a Fermi gas, or Bose-Einstein for a Bose gas. Nevertheless,
the functionals (\ref{eq:cHdef}) and
(\ref{eq:qHdef}) are seemingly different, and in this section we will show that both are
related by the correspondence principle.

>>>>

In sections \ref{sec:classical} and \ref{sec:quantum}, we have seen that the variational
method can be applied to $H$-functionals, which correctly describes the behavior of
classical and quantum dilute gases, as to their respective time-evolution.
Both $H$-functionals defined in Eqs.~(\ref{eq:cHdef}) and
(\ref{eq:qHdef}) also recovers the well-known distribution functions, either Maxwell-Boltzmann
for a classical gas, Fermi-Dirac for a Fermi gas, or Bose-Einstein for a Bose gas. Nevertheless,
the functionals (\ref{eq:cHdef}) and
(\ref{eq:qHdef}) are seemingly different, and in this section we will show that both are
related by the correspondence principle.

### 2021-01-26 tex/relaxproc.tex
Additions to intro.

  \section{Relaxation processes in degenerated quantum gases}\label{sec:relaxproc}
In order to obtain an equation of motion for the quantum gas, we evaluate 
$\Delta H=H(t_2)-H(t_1)$ where the $H$ functional is evaluated in diferent times such that $t_2 > t_1$.
We begin using the following notation $f_{nM}(t_2)\equiv f'_{nM}$ and 
$f_{nM}(t_1) \equiv f''_{nM}$. 
Now, using (\ref{eq:qHdef}) to calculate $\Delta H$ with the previous notation, we obtain
\begin{eqnarray} \label{eq:deltaHdef}
	\Delta H &=& \sum_{M}^{K}\sum_n [f'_{nM} \ln f'_{nM} - f''_{nM}	\ln f''_{nM}\nonumber \\ 
			&&\pm (1\mp f'_{nM}) \ln(1\mp f'_{nM}) \mp (1\mp f''_{nM}) \ln (1\mp f''_{nM})]\delta V.  
\end{eqnarray}
In order to obtain a general equation of motion for the distribution function 
$f_{nM}$, we only take the limit when $(t_2-t_1) \rightarrow 0 $. 
However, our current work is focused on the deviation $g_{nM}$. So, keeping the same workline
we apply the first order approximation (\ref{eq:qFirstOrd}) to (\ref{eq:deltaHdef}) to obtain the following expression
\begin{eqnarray}\label{eq:deltaH1}
	\Delta H &=& \sum_{M}^{K} \sum_n [\bar{f}_n(1+g'_{nM}) \ln \bar{f}_{n}(1+g'_{nM})  -  
			\bar{f}_n(1+g''_{nM}) \ln \bar{f}_n(1+g''_{nM}) \nonumber \\
		    &&  \pm (1 \mp \bar{f}_n \{1+g'_{nM}\}) \ln (1\mp \bar{f}_n\{1+g'_{nM} \})\nonumber \\
		    &&  \mp (1\mp \bar{f}_n \{ 1+g''_{nM} \}) \ln (1\mp \bar{f}_n \{1+g''_{nM} \})]\delta V.\nonumber \\
\end{eqnarray} 
The equation (\ref{eq:deltaH1}) can be rewritten, after some approximations on the logarithm functions, as
\begin{equation}\label{eq:deltaH2}
\Delta H = \sum_{M}^{K}\sum_n [\bar{f}_n(1+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}](g'_{nM}-g''_{nM})\delta V.
\end{equation}
The equation (\ref{eq:deltaH2}) must be analyzed taking the limit $(t_2-t_1) \rightarrow 0$ 
to obtain the equation of motion for the deviation $g_{nM}$.
This analysis will not be treated in this article.

>>>>

\section{Relaxation processes in degenerated quantum gases}\label{sec:relaxproc}

In order to obtain a time-evolution equation for an out-of-equilibrium quantum gas,
we propose the following approach. We start by evaluating $\Delta\mathcal{H}=
\mathcal{H}(t_2)-\mathcal{H}(t_1)$,
where our quantum $H$-functional ---Eq.~(\ref{eq:qHdef})--- is evaluated at diferent times $t_1$ and $t_2$, with
$t_2 > t_1$. This yields:
%
\begin{eqnarray}\label{eq:deltaHdef}
	\Delta\mathcal H &=& \sum_{M=1}^{K}\sum_n [f''_{nM} \ln f''_{nM} - f'_{nM}	\ln f'_{nM}\nonumber \\ 
			&&\pm (1\mp f''_{nM}) \ln(1\mp f''_{nM}) \mp (1\mp f'_{nM}) \ln (1\mp f'_{nM})]\delta V_M. 
\end{eqnarray}
%
In the above equation, and for the rest of this section, we use the short-hand notation
$f_{nM}(t_2)\equiv f''_{nM}$ and 
$f_{nM}(t_1) \equiv f'_{nM}$.
Subsequently, in Eq.~(\ref{eq:deltaHdef}), we replace the distribution functions by their expresions in terms
of deviations from equilibrium ---Eq.~(\ref{eq:qFirstOrd})---, which renders:
\begin{eqnarray}\label{eq:deltaH1}
	\Delta\mathcal H &=& \sum_{M=1}^{K} \sum_n\Big[\bar{f}_n(1+g''_{nM}) \ln \bar{f}_{n}(1+g''_{nM})  -  
			\bar{f}_n(1+g'_{nM}) \ln \bar{f}_n(1+g'_{nM})\Big. \nonumber \\
		    &&  \pm \big(1 \mp \bar{f}_n \{1+g''_{nM}\}\big) \ln \big(1\mp \bar{f}_n\{1+g''_{nM} \}\big)\nonumber \\
		    &&  \mp \big(1\mp \bar{f}_n \{ 1+g'_{nM} \}\big) \ln\big(1\mp \bar{f}_n \{1+g'_{nM} \}\big)\Big.\Big]
		    \delta V_M.
\end{eqnarray}
Afterwards, we expand the logarithmic terms up to the first-order in $g'_{nM}$ and $g''_{nM}$
and rearrange the result, which gives:
%
\begin{equation}\label{eq:deltaH2}
\Delta\mathcal H = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}](g''_{nM}-g'_{nM})\delta V_M.
\end{equation}
%
Finally we divide Eq.~(\ref{eq:deltaH2}) by $\Delta t\equiv t_2-t_1$, and take the limit
$\Delta t\to0$ to obtain:
%
\begin{equation}\label{eq:deltaH3}
\frac{d\mathcal H}{dt} = \sum_{M}^{K}\sum_n [\bar{f}_n(1
	+\ln \bar{f}_n)-\bar{f}_n \{\ln(1\mp \bar{f}_n)\mp 1 \}]\left(\frac{g_{nM}}{dt}\right)\delta V_M.
\end{equation}
%

Eq.~(\ref{eq:deltaH3}) is, within our framework, the time-evolution equation for $g_{nM}$.
Clearly, for describing a concrete situation, it is required to provide a specific approximation for the deviation function $g_{nM}$, however this will be the subject of future work.
